---
title: "Get NASS stats"
author: "jill deines"
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---
  
Goal: retrieve NASS statistics to describe study area. 

Includes:

* county irrigated area - Census
* county total cropland area - Census

* maize production history
* soybean production history

Notes:

* line 57 requires a NASS API key that can be obtained from https://quickstats.nass.usda.gov/api
* data downloads can already be found in the "data/tabular/nass" folder

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      eval = FALSE)
```

**R Packages Needed**
  
```{r packages, warning=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(tidycensus)

# access quickstats API through rnassqs package, using your personal API key
library(rnassqs)
yourKey <- 'enterKeyHere'
nassqs_auth(key = yourKey)

library(here)
sessionInfo()
```


*Directories*
  
```{r directories}
repoDir <- here::here()
nassDir <- paste0(repoDir,'/data/tabular/NASS')

# output filenames: irrigated and total croplands
nassFileName_raw_irrigated <- 'NASS_Census_IrrArea_CornBeltWide_1997-2022.csv'
nassFileName_raw_cropland <- 'NASS_Census_CroplandArea_CornBeltWide_1997-2022.csv'

# output filenames: production
nassFileName_raw_maize_NAME <- 'NASS_Survey_RAW_Maize_production_yields_harvestedArea_CONUS_1997-2023_20250323.csv'

nassFileName_raw_SOYBEAN_NAME <- 'NASS_Survey_RAW_Soybeans_production_yields_harvestedArea_CONUS_1997-2023_20250323.csv'
```


# Irrigated Area
Run 1x

census > economics > farms and land and assets > ag land > area >

Run for all states

```{r retrieveNASS_irr, eval = FALSE}

#statesWanted <- c('IN','IL','IA','OH','MI','MO','MN','WI','MI','NE','SD','ND')

# param list without state
paramsNoState <- list("source_desc" = "CENSUS",
                 "agg_level_desc" = 'COUNTY',
                 "year__GE" = 1997,
                 "short_desc" = 'AG LAND, IRRIGATED - ACRES',
                 "domain_desc" = 'TOTAL')


# do all states together
nassdf <- nassqs(paramsNoState)


# # split up by state to avoid query limits
# nasslist <- list()
# for(state in statesWanted){
#   params <- c(list("state_alpha" = state), paramsNoState)
#   # retrieve county data by state
#   nasslist[[state]] <- nassqs(params)
# }

# # convert list of data frames to 1 giant dataframe
# nass.df <- do.call("rbind",nasslist)

# tidy up a bit
nassIrr <- nassdf %>%
  mutate(fips5 = paste0(state_fips_code, county_ansi)) %>%
  filter(!is.na(Value)) %>%    # remove no data
  mutate(irrArea_ha = Value / 2.471) %>%  # acres to ha
  dplyr::rename(irrArea_acres = Value) %>%
  dplyr::select(c(fips5, county_name, state_alpha, state_name, year, short_desc,
                  domain_desc, irrArea_acres, irrArea_ha)) %>%
  arrange(fips5, year) %>%
  as_tibble()

# export
write_csv(nassIrr, paste0(nassDir, '/', nassFileName_raw_irrigated))

```



# County Cropland Area
census > economics > farms and land and assets > ag land > area >

```{r getCropArea, eval = FALSE}


# param list without state
paramsNoState2 <- list("source_desc" = "CENSUS",
                 "agg_level_desc" = 'COUNTY',
                 "year__GE" = 1997,
                 "short_desc" = 'AG LAND, CROPLAND - ACRES',
                 "domain_desc" = 'TOTAL')

# all at once
nass.df <- nassqs(paramsNoState2)


# tidy up a bit
nassCropland <- nass.df %>%
  mutate(fips5 = paste0(state_fips_code, county_ansi)) %>%
  filter(!is.na(Value)) %>%    # remove no data
  mutate(croplandArea_ha = Value / 2.471) %>%  # acres to ha
  dplyr::rename(croplandArea_acres = Value) %>%
  dplyr::select(c(fips5, county_name, state_alpha, state_name, year, short_desc,
                  domain_desc, croplandArea_ha, croplandArea_acres)) %>%
  arrange(fips5, year) %>%
  as_tibble()

# export
write_csv(nassCropland, paste0(nassDir, '/', nassFileName_raw_cropland))

```












# Corn metrics
Run 1x

survey > crops > field crops

Note: this ran with year >= 1997 in 2023, but it fails now in 2026 due to too many records. Would need to adjust years or download by state, etc

```{r retrieveNASS_corn, eval = FALSE}

# string of "short_desc" wanted
short_corn <- c('CORN, GRAIN - PRODUCTION, MEASURED IN BU',
                'CORN, GRAIN, IRRIGATED - PRODUCTION, MEASURED IN BU',
                'CORN, GRAIN, NON-IRRIGATED - PRODUCTION, MEASURED IN BU',
                'CORN, GRAIN - YIELD, MEASURED IN BU / ACRE',
                'CORN, GRAIN, IRRIGATED - YIELD, MEASURED IN BU / ACRE',
                'CORN, GRAIN, NON-IRRIGATED - YIELD, MEASURED IN BU / ACRE',
                'CORN, GRAIN - ACRES HARVESTED',
                'CORN, GRAIN, IRRIGATED - ACRES HARVESTED',
                'CORN, GRAIN, NON-IRRIGATED - ACRES HARVESTED')

# param list without short desc
paramsNoState <- list("source_desc" = "SURVEY",
                 "agg_level_desc" = 'COUNTY',
                 "year__GE" = 1997,
                 "domain_desc" = 'TOTAL')


# split up by short-desc to avoid query limits, etc
nass_corn_ls <- list()
for(short_desc in short_corn){
  params <- c(list("short_desc" = short_desc), paramsNoState)
  # retrieve county data by state
  nass_corn_ls[[short_desc]] <- nassqs(params)
}

# # convert list of data frames to 1 giant dataframe
nass_corn <- do.call("rbind",nass_corn_ls)

# export
write_csv(nass_corn, paste0(nassDir, '/', nassFileName_raw_maize_NAME))

```

# Soy metrics
Run 1x

survey > crops > field crops

```{r retrieveNASS_soy, eval = FALSE}

# string of "short_desc" wanted
short_SOY <- c('SOYBEANS - PRODUCTION, MEASURED IN BU',
                'SOYBEANS, IRRIGATED - PRODUCTION, MEASURED IN BU',
                'SOYBEANS, NON-IRRIGATED - PRODUCTION, MEASURED IN BU',
                'SOYBEANS - YIELD, MEASURED IN BU / ACRE',
                'SOYBEANS, IRRIGATED - YIELD, MEASURED IN BU / ACRE',
                'SOYBEANS, NON-IRRIGATED - YIELD, MEASURED IN BU / ACRE',
                'SOYBEANS - ACRES HARVESTED',
                'SOYBEANS, IRRIGATED - ACRES HARVESTED',
                'SOYBEANS, NON-IRRIGATED - ACRES HARVESTED')

# param list without short desc
paramsNoState <- list("source_desc" = "SURVEY",
                 "agg_level_desc" = 'COUNTY',
                 "year__GE" = 1997,
                 "domain_desc" = 'TOTAL')


# split up by short-desc to avoid query limits, etc
nass_soy_ls <- list()
for(short_desc in short_SOY){
  params <- c(list("short_desc" = short_desc), paramsNoState)
  # retrieve county data by state
  nass_soy_ls[[short_desc]] <- nassqs(params)
}

# # convert list of data frames to 1 giant dataframe
nass_soy <- do.call("rbind",nass_soy_ls)

# export
write_csv(nass_soy, paste0(nassDir, '/', nassFileName_raw_SOYBEAN_NAME))

```

