---
title: "Check NASS - Irrigation by Crop"
author: "jill deines"
date: '2025-05-13'
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---
  
Goal: look at irrigated area by crop type from NASS statistics

output: formmated data frame for figure 2a


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path='../figure/00.30_checkNass_irrStats/',
                      cache = FALSE,
                      eval = FALSE)
```

**R Packages Needed**
  
```{r packages, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(tidycensus)
library(purrr)

library(data.table)


library(here)
sessionInfo()
```


*Directories*
  
```{r directories}
# full 2017 census to extract irrigation by crop type
census2017raw_fn <- '/Users/dein121/local/data_nonRepo/2025_quickstats/qs.census2017.txt.gz'


repoDir <- here::here()
nassDir <- paste0(repoDir,'/data/tabular/NASS')
nassFileName_raw_cropland <- 'NASS_Census_CroplandArea_CornBeltWide_1997-2022.csv'

# output
compiled_nassIrrDataName <- 'Compiled_CENSUS_2017_irrigated_byCropType_7states.csv'
```


# load and clean data

## extract info from census
works from raw 2017 census to parse through crop types and suprasets

```{r load1, eval = FALSE}

states_toKeep <- c('MN','WI','MI','OH','IN','IL','IA')

columnsToDrop <- c('STATE_ANSI','ASD_CODE','ASD_DESC','COUNTY_ANSI','COUNTY_CODE',
                   'COUNTY_NAME','REGION_DESC','ZIP_5','WATERSHED_CODE',
                   'WATERSHED_DESC','CONGR_DISTRICT_CODE','COUNTRY_CODE',
                   'COUNTRY_NAME','LOCATION_DESC','FREQ_DESC','BEGIN_CODE',
                   'END_CODE','REFERENCE_PERIOD_DESC','WEEK_ENDING','LOAD_TIME',
                   'CV_.','STATE_NAME','STATE_FIPS_CODE')


# load and format/clean
test <- read.table(census2017raw_fn, sep = '\t', header = TRUE, fill = TRUE)

# remove (D) values, remove commas (!), convert to numeric
nass.df <- test[!grepl('(D)',test$VALUE),]
nass.df$VALUE <- gsub(",", "", nass.df$VALUE, fixed = T)
nass.df$VALUE <- as.numeric(nass.df$VALUE)

test2 <- nass.df %>% 
  filter(!is.na(nass.df$VALUE)) %>%
  dplyr::select(-all_of(columnsToDrop)) %>%
  filter(STATE_ALPHA %in% states_toKeep)

test3 <- test2 %>% 
  filter(SECTOR_DESC %in% c('CROPS')) %>%
  filter(GROUP_DESC %in% c('FIELD CROPS','VEGETABLES','FRUIT & TREE NUTS',
                           'HORTICULTURE')) %>%
  filter(grepl('IRRIGATED',PRODN_PRACTICE_DESC)) %>%
  filter(AGG_LEVEL_DESC == 'STATE',
         UNIT_DESC == 'ACRES',
         DOMAIN_DESC == 'TOTAL',
         !grepl('NONE OF CROP',PRODN_PRACTICE_DESC),
         !grepl('NON-IRRIGATED',PRODN_PRACTICE_DESC)) %>%
  mutate(values_ha = VALUE * 0.404686) %>%
  dplyr::select(-c(AGG_LEVEL_DESC,UNIT_DESC, DOMAIN_DESC,
                   SOURCE_DESC, GROUP_DESC, SECTOR_DESC,
                   UTIL_PRACTICE_DESC, STATISTICCAT_DESC, DOMAINCAT_DESC,
                   VALUE)) %>%
  dplyr::select(YEAR, STATE_ALPHA, COMMODITY_DESC, everything()) %>%
  arrange(YEAR, STATE_ALPHA, COMMODITY_DESC) %>%
  mutate(YEAR = as.numeric(YEAR))


# collapse hay
hays0 <- test3 %>%
  filter(grepl('HAY',COMMODITY_DESC))
hays <- hays0 %>%
  filter(COMMODITY_DESC == 'HAY & HAYLAGE',
         CLASS_DESC == 'ALL CLASSES',
         PRODN_PRACTICE_DESC == 'IRRIGATED')

# collapse berries (no total in 2002)
berries0 <- test3 %>%
  filter(grepl('BERRY',COMMODITY_DESC))
berries <- berries0 %>%
  group_by(YEAR, STATE_ALPHA, COMMODITY_DESC, CLASS_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  mutate(PRODN_PRACTICE_DESC = 'IRRIGATED',
         SHORT_DESC = 'BERRY TOTALS, IRRIGATED - ACRES HARVESTED')

# ALL OTHERS
others0 <- test3 %>%
    filter(!grepl('BERRY',COMMODITY_DESC),
           !grepl('HAY',COMMODITY_DESC),
           CLASS_DESC != 'WINTER',
           !grepl('SPRING',CLASS_DESC),
           !grepl('PEPPERMINT',CLASS_DESC),
           !grepl('SPEARMINT',CLASS_DESC),
           !grepl('OIL TYPE',CLASS_DESC),
           !grepl('PART OF CROP', PRODN_PRACTICE_DESC),
           !grepl('ENTIRE CROP',PRODN_PRACTICE_DESC))

# collapse corn + sorghum grain/silage
others <- others0 %>%
  group_by(YEAR, STATE_ALPHA, COMMODITY_DESC, CLASS_DESC,
           PRODN_PRACTICE_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  mutate(SHORT_DESC = paste0(COMMODITY_DESC, ', IRRIGATED - ACRES HARVESTED'))


# combine 
cen2017 <- bind_rows(hays, berries, others) %>%
  arrange(YEAR, STATE_ALPHA, COMMODITY_DESC)
```



## part 1: irrigated area by crop type
```{r irr_byCrop}

# all study area states together
latest_totals <- cen2017 %>%
  group_by(COMMODITY_DESC) %>%
  summarise(values_ha = sum(values_ha, na.rm = TRUE)) %>%
  ungroup()

# create and add others class for low freq types
other <- latest_totals %>%
  filter(values_ha <=5000) %>%
  summarize(values_ha = sum(values_ha)) %>%
  mutate(COMMODITY_DESC = 'OTHER')
  
latest_totals2 <- latest_totals %>%
  filter(values_ha > 5000) %>%
  bind_rows(other)

# make a percent column
total_irr <- sum(latest_totals2$values_ha)
latest_totals2$percentOfIrr <- latest_totals2$values_ha / total_irr * 100
latest_totals2 %>% arrange(-percentOfIrr)

# plot! by area
p_irrByCrop <- ggplot(latest_totals2 %>% filter(values_ha > 5000),
       aes(x = reorder(COMMODITY_DESC, -values_ha), y = values_ha / 1000)) +
  geom_bar(stat = 'identity') +
  xlab('Commodity') + ylab('Area (1000 ha)') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 


# percent
# plot!
irrArea_toPlot <- latest_totals2 %>% filter(values_ha > 5000)
p_irrByCrop_percent <- ggplot(irrArea_toPlot,
       aes(x = reorder(COMMODITY_DESC, -percentOfIrr), y = percentOfIrr )) +
  geom_bar(stat = 'identity') +
  xlab('Commodity') + ylab('Percent of Irrigated Area') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

## part 2: get croptype areas, total 
percent of cropped area covered by each crop group

also works from full 2017 census. some cleaning to remove duplicates via suprasets. this is not particularly efficient but done iteratively while examining subgroups

```{r stats_rainfedProportions}
states_toKeep <- c('MN','WI','MI','OH','IN','IL','IA')

# total cropland
crop_county <- read_csv(paste0(nassDir,'/',nassFileName_raw_cropland))
crop_state <- crop_county %>%
  group_by(state_alpha, state_name, year) %>%
  summarize(croplandArea_ha = sum(croplandArea_ha)) %>%
  ungroup()

cropland_region <- crop_state %>% 
  filter(state_alpha %in% states_toKeep,
         year == 2017) 
croplandTotal <- sum(cropland_region$croplandArea_ha)

# from formatted 2017 census (previous chunk) extract corn/soy total area in study region
corn_soy <- nass.df %>% 
  filter(!is.na(nass.df$VALUE)) %>%
  filter(STATE_ALPHA %in% states_toKeep) %>%
  filter(PRODN_PRACTICE_DESC == 'ALL PRODUCTION PRACTICES',
         grepl('CORN, GRAIN', SHORT_DESC) | grepl('SOY', SHORT_DESC)) %>%
  filter(AGG_LEVEL_DESC == 'STATE',
         UNIT_DESC == 'ACRES',
         DOMAIN_DESC == 'TOTAL') %>%
  mutate(values_ha = VALUE * 0.404686) %>%
  group_by(SHORT_DESC) %>%
  summarise(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  mutate(percentCropland = values_ha / croplandTotal * 100,
         commodity = case_when(grepl('CORN',SHORT_DESC) ~ 'CORN',
                               grepl('SOY',SHORT_DESC) ~ 'SOYBEANS'))
corn_soy


# extract other groups in Figure 2a -----------
nass2017_a <- nass.df %>% 
  filter(!is.na(nass.df$VALUE)) %>%
  filter(STATE_ALPHA %in% states_toKeep) %>%
  filter(PRODN_PRACTICE_DESC == 'ALL PRODUCTION PRACTICES',
         SECTOR_DESC == 'CROPS',
         AGG_LEVEL_DESC == 'STATE',
         UNIT_DESC == 'ACRES',
         DOMAIN_DESC == 'TOTAL') # RETAIN hay and haylage combo column

vegetables <- nass2017_a %>%
  filter(GROUP_DESC == 'VEGETABLES') %>%
  dplyr::select(c(COMMODITY_DESC, CLASS_DESC, 
                  SHORT_DESC,STATE_ALPHA, VALUE)) %>%
  mutate(values_ha = VALUE * 0.404686) %>%
  arrange(COMMODITY_DESC) %>%
  filter(!(grepl('FRESH MARKET', SHORT_DESC) | 
             grepl('PROCESSING', SHORT_DESC) |
             grepl('SUMMER', SHORT_DESC) |
             grepl('WINTER', SHORT_DESC) |
             grepl('BEANS',SHORT_DESC) |
             grepl('CORN',SHORT_DESC))) %>%
  group_by(SHORT_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  summarize(values_ha = sum(values_ha)) %>%
  mutate(commodity = 'VEGETABLE TOTALS',
         percentCropland = values_ha / croplandTotal * 100)
                  
hays <- nass2017_a %>%
  filter(SHORT_DESC == 'HAY - ACRES HARVESTED') %>%
  mutate(values_ha = VALUE * 0.404686) %>%
  group_by(SHORT_DESC,COMMODITY_DESC,CLASS_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  arrange(COMMODITY_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  mutate(commodity = 'HAY & HAYLAGE',
         percentCropland = values_ha / croplandTotal * 100)

berries <- nass2017_a %>%
  filter(SHORT_DESC == 'BERRY TOTALS - ACRES GROWN') %>%  
  mutate(values_ha = VALUE * 0.404686) %>%
  group_by(SHORT_DESC,COMMODITY_DESC,CLASS_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  arrange(COMMODITY_DESC) %>%
  mutate(commodity = 'BERRY TOTALS',
         percentCropland = values_ha / croplandTotal * 100)
  
beans <- nass2017_a %>%
  filter(grepl('BEANS',SHORT_DESC)) %>%  
  mutate(values_ha = VALUE * 0.404686) %>%
  group_by(SHORT_DESC,COMMODITY_DESC,CLASS_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  arrange(COMMODITY_DESC) %>%
  filter(!(grepl('FRESH MARKET', SHORT_DESC) | 
             grepl('PROCESSING', SHORT_DESC) |
             grepl('SOY',SHORT_DESC))) %>%
  summarize(values_ha = sum(values_ha)) %>%
  mutate(commodity = 'BEANS',
         percentCropland = values_ha / croplandTotal * 100)

wheat <- nass2017_a %>%
  filter(SHORT_DESC ==  'WHEAT, SPRING, (EXCL DURUM) - ACRES HARVESTED' |
           SHORT_DESC == 'WHEAT, WINTER - ACRES HARVESTED') %>%  
  filter(grepl('WHEAT',SHORT_DESC)) %>%
  mutate(values_ha = VALUE * 0.404686) %>%
  group_by(SHORT_DESC,COMMODITY_DESC,CLASS_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  arrange(COMMODITY_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  mutate(commodity = 'WHEAT',
         percentCropland = values_ha / croplandTotal * 100)

# wild rice
wildRice <- nass2017_a %>%
  filter(grepl('RICE',SHORT_DESC)) %>%  
  mutate(values_ha = VALUE * 0.404686) %>%
  group_by(SHORT_DESC,COMMODITY_DESC,CLASS_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  arrange(COMMODITY_DESC) %>%
  mutate(commodity = 'WILD RICE',
         percentCropland = values_ha / croplandTotal * 100)

# popcorn
popcorn <- nass2017_a %>%
  filter(grepl('POPCORN',SHORT_DESC)) %>%  
  mutate(values_ha = VALUE * 0.404686) %>%
  group_by(SHORT_DESC,COMMODITY_DESC,CLASS_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  arrange(COMMODITY_DESC) %>%
  mutate(commodity = 'POPCORN',
         percentCropland = values_ha / croplandTotal * 100)

orchards <- nass2017_a %>%
  filter(GROUP_DESC == 'FRUIT & TREE NUTS',
         !(grepl('BERRIES',SHORT_DESC ) | grepl('BERRY',SHORT_DESC)),
         # eliminate bearing/non bearing sub cats
         grepl('&', SHORT_DESC) | grepl('GROWN',SHORT_DESC),
         # REMOVE OTHER REPETITIVE SUB CATS
         !(CLASS_DESC %in% c('CLINGSTONE','FREESTONE','(EXCL BARTLETT)',
                             'BARTLETT','IMPROVED','NATIVE & SEEDLING')),
         # keep plum and prune indvl classes
         COMMODITY_DESC != 'PLUMS & PRUNES',
         COMMODITY_DESC != 'TREE NUT TOTALS') %>% 
  mutate(values_ha = VALUE * 0.404686) %>%
  group_by(SHORT_DESC,COMMODITY_DESC,CLASS_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  ungroup() %>%
  arrange(COMMODITY_DESC) %>%
  summarize(values_ha = sum(values_ha)) %>%
  mutate(commodity = 'ORCHARDS',
         percentCropland = values_ha / croplandTotal * 100)

other <- data.frame(values_ha = 13360,
                    commodity = 'OTHER',
                    percentCropland = 13360 / croplandTotal * 100)
# combine and derive "other"
allCroplandByCrop <- bind_rows(corn_soy,vegetables,hays,berries,beans,
                               wheat,wildRice,popcorn, orchards, other)

# p_croplandByCrop_percent <- ggplot(allCroplandByCrop,
#        aes(x = reorder(commodity, -percentCropland), y = percentCropland )) +
#   geom_bar(stat = 'identity') +
#   xlab('Commodity') + ylab('Percent of Cropland Area') +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
# 
# p_croplandByCrop_percent 

```


## combine and export
```{r}

comboCrop <- irrArea_toPlot %>%
  mutate(type = 'Irrigated Land') %>% 
  rename(percent = percentOfIrr) %>%
  bind_rows(allCroplandByCrop %>%
              dplyr::select(commodity, values_ha, percentCropland) %>%
              rename(COMMODITY_DESC = commodity,
                     percent = percentCropland) %>%
              mutate(type = 'All Cropland')) 
comboCrop$type2 <- factor(comboCrop$type, levels=c('Irrigated Land',
                                                   'All Cropland'))

write_csv(comboCrop, paste0(nassDir,'/',compiled_nassIrrDataName))

```