---
title: "Yield Impacts: Causal Impacts"
author: "jill deines"
date: '2022-11-21'
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---
  
  
Goal: Sensitivity analysis (for field size) for maize yield impact evaluation. This replicates the maize causal forest analysis based on an independent field sample using a field size threshold of 100 pixels

* input master data file available in associated Zenodo repo
* intermediate causal forest objects available in associated Zenodo repo


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path='../figure/03.02_yieldImpacts_SA/',
                      cache = FALSE)
```

**R Packages Needed**
  
```{r packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(grf)
library(earth)
library(corrplot)
library(RColorBrewer)

library(here)
sessionInfo()
```


*Directories*
  
```{r directories}
repoDir <- here::here()

# master data
dataDir <- '/Users/dein121/local/data_nonRepo/2025_irrigationAndYields/final_data_repo'

dataName <- 'pointSample_combined_fewerCDL_test100px_20260123.rds'

# scratch folder for model rdata objects
scratchFolder <- paste0(dataDir,'/causal_forest_objects/maize_100px')

# make scratch folder if necessary
dir.create(file.path(scratchFolder), showWarnings = FALSE)

```



# Load and filter
Clean master data by:

* maize only
* remove spurious classifications (eg, annual status differs from overarching irrigation type)
* clean extreme yield values


```{r loadMaster}
# load data, extract corn
master <- readRDS(paste0(dataDir,'/',dataName)) %>%
  # drop columns relating to climate projections
  dplyr::select(-c(contains('vpd_summ')))


maize <- master %>% 
  filter(cdl_class == 'Corn',
         !is.na(yield_tha_corn))
nrow(maize)

# assign treatment variable, W (1 = irrigated that year, 0 is not)
maize2 <- maize %>%
  mutate(W = case_when(lanid_status == 1 ~ 1,
                       is.na(lanid_status) ~ 0))

# check status distribution
table(maize2$irr_type, maize2$W)

# remove likely bad single-year classifications
maize3 <- maize2 %>%
  filter(!(irr_type == 'never - random' & W == 1),
         !(irr_type == 'existing' & W == 0),
         !(irr_type == 'new' & W == 0)) # probably prior to adoption but still 

# check status distribution
table(maize3$irr_type, maize3$W)

# filter for yield data and remove extreme yield values
cleanQ <- quantile(maize3$yield_tha_corn, probs = c(0.0001, 0.9999), na.rm=TRUE)
cleanQ   

maize4 <- maize3 %>%
  filter(yield_tha_corn > cleanQ[1]) %>%
  filter(yield_tha_corn < cleanQ[2]) %>%
  mutate(uniqueID = row_number()) %>%
  dplyr::rename(., Y = yield_tha_corn)          # set outcome variable

# breakdown of treatment/control
table(maize4$W)
table(maize4$year)
table(maize4$W, maize4$year)

```

# Causal Forest

## Propensity model

```{r propensity, eval = FALSE}

X_norm <- maize4 %>%
  dplyr::select(c(contains('norm'), contains('dist'), 'wtd_zell',
                  "nccpicorn",'drclasscd',
                  'clay','sand','silt',
                  'rootznaws', 'slope','bulkDensityt','ksat','pH')) %>%
  dplyr::select(-c(dist_toIrr))

names(X_norm)

W <- maize4 %>% dplyr::pull(W)

W_forest <- grf::regression_forest(X_norm, W, num.trees = 2000)

W_hat <- predict(W_forest)$predictions

WvarImp <- W_forest %>% 
  grf::variable_importance() %>% 
  as.data.frame() %>% 
  mutate(variable = colnames(W_forest$X.orig)) %>% 
  arrange(desc(V1))

# save intermediates
saveRDS(maize4, paste0(scratchFolder, '/maize.rds'))
saveRDS(W_forest, paste0(scratchFolder, '/W_forest.rds'))
saveRDS(W_hat, paste0(scratchFolder, '/W_hat.rds'))
saveRDS(WvarImp, paste0(scratchFolder, '/W_forest_varImp.rds'))
```

## Prune samples

toss out rows with excessively high/low propensities

```{r prune, eval = FALSE}
W <- maize4 %>% dplyr::pull(W)

W_hat <- readRDS(paste0(scratchFolder, '/W_hat.rds'))
WvarImp <- readRDS(paste0(scratchFolder, '/W_forest_varImp.rds'))
WvarImp

# check propensity scores
p_rf = W_hat
hist(p_rf, xlab = 'Propensity Scores', main = 'Histogram of Propensity Scores')

sum(p_rf < .05)
sum(p_rf > .95)

{plot(smooth.spline(p_rf, W, df = 4))
abline(0, 1)}

# filter it
overlap0 <- maize4 %>%
  mutate(w_hat = W_hat) %>%
  mutate(keep = case_when(w_hat < 0.02 | w_hat > 0.98 ~ 0,
                        w_hat >= 0.02 | w_hat <= 0.98  ~ 1))

overlap <- overlap0 %>% filter(keep == 1)

nrow(overlap0) - nrow(overlap)

nrow(overlap)

both <- data.frame(W_hat = W_hat,
                   type = paste0('original sample (',nrow(maize4), ' obs)')) %>%
  bind_rows(data.frame(W_hat = overlap$w_hat,
                       type = paste0('Propensity Filtered (', nrow(overlap), ' obs)')))

ggplot(both,
       aes(x = W_hat)) +
  geom_histogram() +
  geom_vline(xintercept = 0.02, col='red') +
  geom_vline(xintercept = 0.98, col = 'red') +
  facet_wrap(~type, nrow = 2) +
  #ggeasy::easy_center_title() +
  ggtitle('Maize') +
  theme_bw()


table(overlap$W)

table(overlap$year)

table(overlap[,c('state')])

table(overlap[,c('state',"W")])

# save overlap data!
saveRDS(overlap, paste0(scratchFolder, '/overlap.rds'))
```

## Expected Outcome: yhat

### pick covars

Here, removes gldas variables (redundant with tc soilm) and tc_def (redundant with vpd)

```{r earth, eval = FALSE}
overlap <- readRDS(paste0(scratchFolder, '/overlap.rds'))

# subset of variables to choose from
X_c1 <- overlap %>%
  dplyr::select(-c (W,  geom_id, fips5, w_hat, keep, cdl_class,contains('norm'),
                    latitude, longitude, mukey, state_fips, irr_type, uniqueID, 
                    state_name, state,  nccpisoy, yield_tha_soy, lanid_status,
                    adoptionYear, contains('dist'), elevation, ted, 
                    lastYearIrr, contains('condon'),
                    contains('gldas'), contains('tc_def'))) %>%
  tidyr::drop_na()

# mars for var dplyr::selection
set.seed(5)
marsAll <- earth::earth(Y ~ ., data = X_c1, degree = 2)

summary(marsAll)
ev <- evimp(marsAll) # estimate variable importance
ev

saveRDS(marsAll, paste0(scratchFolder, '/mars_c1.rds'))
```

### Train Yhat

```{r Yhat, eval = FALSE}

# variables from mars
marsAll <- readRDS(paste0(scratchFolder, '/mars_c1.rds'))
summary(marsAll)
ev <- evimp(marsAll) # estimate variable importance
ev

varsToKeep = rownames(ev)
X <- overlap %>%
  dplyr::select(all_of(varsToKeep)) 

Y <- overlap %>% dplyr::pull(Y)
Y_forest <- grf::regression_forest(X, Y, num.trees = 2000)

YvarImp <- Y_forest %>% 
  grf::variable_importance() %>% 
  as.data.frame() %>% 
  mutate(variable = colnames(Y_forest$X.orig)) %>% 
  arrange(desc(V1))
YvarImp

Y_hat <- predict(Y_forest)$predictions

saveRDS(Y_hat, paste0(scratchFolder, '/Y_hat.rds'))
saveRDS(Y_forest, paste0(scratchFolder, '/Y_forest.rds'))
saveRDS(YvarImp, paste0(scratchFolder, '/Y_forest_varImp.rds'))
```

### evaluate outcome model

```{r outcomeEval, eval = FALSE}
overlap <- readRDS(paste0(scratchFolder, '/overlap.rds'))
Y_hat <- readRDS(paste0(scratchFolder, '/Y_hat.rds'))
YvarImp <- readRDS(paste0(scratchFolder, '/Y_forest_varImp.rds'))
YvarImp

Y <- overlap %>% dplyr::pull(Y)

### Outcome Estimation Y_Hat
Ychecker <- data.frame(Y = Y,
                       Y_hat = Y_hat,
                       year = overlap$year)
Yagreement <- lm(Y~Y_hat)
summary(Yagreement)

# yield year by year
YcheckerLong <- Ychecker %>%
  tidyr::gather(., key = variable, value = value, Y:Y_hat)

ggplot(YcheckerLong, aes(x=as.factor(year), y = value, group = interaction(year,variable), 
                         fill = variable)) +
  geom_boxplot() +
  theme_bw() +
  ylab('Yield (t/ha)') +
  xlab('Year')

ggplot(Ychecker, aes(x=Y, y = Y_hat)) +
  geom_hex() +
  scale_fill_gradientn(colors = brewer.pal(9,'YlGnBu')) +
  geom_smooth(method= 'lm') +
 # coord_equal(xlim=c(1,3), ylim=c(1,3)) + 
  geom_abline(slope = 1, intercept = 0, linetype='dashed') + 
  facet_wrap(~year) +
  theme_bw()
```

## causal forest

```{r causal, eval = FALSE}
# reload overlap
overlap <- readRDS(paste0(scratchFolder, '/overlap.rds'))
Y_hat <- readRDS(paste0(scratchFolder, '/Y_hat.rds'))

# reload mars variables
marsAll <- readRDS(paste0(scratchFolder, '/mars_c1.rds'))
ev <- evimp(marsAll) # estimate variable importance
envvars <- rownames(ev)

# treatment
W <- overlap %>% dplyr::pull(W)
w_hat <- overlap$w_hat
#WvarImp

# outcome
Y <- overlap %>% dplyr::pull(Y)

# make covariate list; 
# add in important propensity variables not already present
varsToKeep_main = c(envvars, 'rootznaws','wtd_zell' )


Xmain <- overlap %>%
  dplyr::select(all_of(varsToKeep_main) )

cf <- grf::causal_forest(Xmain, Y, W, Y_hat, w_hat, num.trees = 2000)

varImpMain <- cf %>% 
  variable_importance() %>% 
  as.data.frame() %>% 
  mutate(variable = colnames(cf$X.orig)) %>% 
  arrange(desc(V1))
varImpMain

saveRDS(cf, paste0(scratchFolder, '/cf_clusters_xc1.rds'))
saveRDS(varImpMain, paste0(scratchFolder, '/cf_clusters_xc1_varImp.rds'))
```


# Checking things out

reload and extract model outputs

```{r reload}
# data used
overlap <- readRDS(paste0(scratchFolder, '/overlap.rds'))

# propensity model results
W_hat <- overlap %>% dplyr::pull(w_hat)
WvarImp <- readRDS(paste0(scratchFolder, '/W_forest_varImp.rds'))
W_hat_all <- readRDS(paste0(scratchFolder, '/W_hat.rds'))

# outcome model
Y_hat <- readRDS(paste0(scratchFolder, '/Y_hat.rds'))
YvarImp <- readRDS(paste0(scratchFolder, '/Y_forest_varImp.rds'))

# causal forest
cf <- readRDS(paste0(scratchFolder, '/cf_clusters_xc1.rds'))
varImp <- readRDS(paste0(scratchFolder, '/cf_clusters_xc1_varImp.rds'))

# reconstruct subsets
X_norm <- maize4 %>%
  dplyr::select(c(contains('norm'), contains('dist'), contains('wtd'),
                  "nccpicorn",'drclasscd',
                  'clay','sand','silt',
                  'rootznaws', 'slope','bulkDensityt','ksat','pH')) %>%
  dplyr::select(-c(dist_toIrr))

W <- overlap %>% dplyr::pull(W)
Y <- overlap %>% dplyr::pull(Y)

# process results ------------------------------
# add predicted treatment effects to data 
constOob_predict <- predict(cf)

# add hats to original data frame
overlapW0 <- overlap %>%
  bind_cols(data.frame(Y_hat = Y_hat))

colNormals <- names(X_norm)

colVars <- varImp %>% dplyr::pull(variable)

overlapW <- overlapW0 %>%
  mutate_at(colNormals, list(Q4n = ~ntile(., 4),
                             Q5n = ~ntile(., 5))) %>%
  mutate_at(colVars, list(Q4 = ~ntile(., 4),
                          Q5 = ~ntile(., 5)))

overlapTau <- overlapW %>%
  bind_cols(constOob_predict) 
```


## Propensity by covariates

```{r propensity_trends, fig.width = 7, fig.height = 6.5, dpi = 300,eval = FALSE}
# add hats to original data frame
prop_all <- maize4 %>%
  bind_cols(data.frame(W_hat = W_hat_all))

# calculate quantiles of interest
colNormals <- names(X_norm)
colVars <- varImp %>% dplyr::pull(variable)

prop_all2 <- prop_all %>%
  mutate_at(WvarImp$variable, list(Q4n = ~ntile(., 4),
                           Q5n = ~ntile(., 5))) %>%
  mutate_at(colVars, list(Q4 = ~ntile(., 4),
                           Q5 = ~ntile(., 5)))

# plot normals vs What : quantiles
propensityVarsLong2 <- prop_all2 %>%
  dplyr::select(c(W_hat, contains('Q4n'))) %>%
  tidyr::gather(., key = variable, value = value, ksat_Q4n:bulkDensityt_Q4n) 

ggplot(propensityVarsLong2,
       aes(y = W_hat, x = value, group = value)) +
  geom_boxplot() +
  facet_wrap(~variable) +
  theme_bw() + xlab('Quartile') + ylab('W.hat = conditional prob of treatment')
```

### propensity fewer

```{r propensityFewer, eval = FALSE}
WvarImp

vars3 <- c('ksat','rootznaws','wtd_zell',
                         'vpd_jul_norm_hPa', 'sand','slope',
                         'clay','silt',
                         'pr_jul_norm_mm')
vars3_q4 <- paste0(vars3, '_Q4n')
propensity2 <- propensityVarsLong2 %>% 
  dplyr::filter(variable %in% vars3_q4)


ggplot(propensity2,
       aes(y = W_hat, x = value, group = value)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = 'free_x') +
  theme_bw() +# xlab('Qaurtile') +
  ylab('W.hat = conditional prob of treatment')
```

# more



## Average treatment Effects

```{r ate}
varImp

ate_cf_aipw = average_treatment_effect(cf, target.sample = 'overlap')

tauhat_rf_aipw = c(ATE=ate_cf_aipw["estimate"],
                   lower_ci=ate_cf_aipw["estimate"] - 1.96 * ate_cf_aipw["std.err"],
                   upper_ci=ate_cf_aipw["estimate"] + 1.96 * ate_cf_aipw["std.err"])

tauhat_rf_aipw

ggplot(overlapTau,
       aes(x = predictions)) +
  geom_histogram() +
  xlab('CATE (t/ha)') +
  geom_vline(xintercept = 0, col = 'black', linetype = 'dashed') +
  geom_vline(xintercept = ate_cf_aipw["estimate"], col = 'red') +
  theme_bw()

summary(overlapTau$Y)
summary(overlapTau$predictions)
quantile(overlapTau$predictions, c(.1, .5, .9)) 
```

### ATE as percent

```{r ate_percent}
# distribution of ATE
summary(overlapTau$predictions)

# ate as a percent of mean  yield
ate_cf_aipw["estimate"]/mean(overlapTau$Y) *100

# ate as the median/mean of the percent increase for each sample point
percentCheck <- overlapTau %>%
  mutate(yield_diff_perc = predictions/Y * 100)

summary(percentCheck$yield_diff_perc)
quantile(percentCheck$yield_diff_perc, c(.1, .5, .9)) 


meanATEp <- mean(percentCheck$yield_diff_perc)
meanATEp

ggplot(percentCheck %>% filter(yield_diff_perc < 100 & yield_diff_perc > -25),
       aes(x = yield_diff_perc)) +
  geom_histogram() +
  xlab('CATE (% of yield)') +
  geom_vline(xintercept = 0, col = 'black', linetype = 'dashed') +
  geom_vline(xintercept = meanATEp, col = 'red') +
  theme_bw()


# save df for official figures
dataOut <- percentCheck %>%
  dplyr::select(c(geom_id, uniqueID, irr_type, fips5, adoptionYear,
                  state_name, year, cdl_class, Y, W, 
                  w_hat, Y_hat,
                  predictions, yield_diff_perc,
                  all_of(YvarImp$variable),
                  all_of(WvarImp$variable)))

#write_csv(dataOut, paste0(dirOut,'/', dfOut))
```




