---
title: "Irrigation likelihood"
author: "jill deines"
date: '2022-12-04'
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---
  
Goal: Assess factors that might predict irrigation installation

Notes:

* creates Figure 4
* input master data file available in associated Zenodo repo


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path='../figure/04.00_irrLikelihood/',
                      cache = FALSE)
```

**R Packages Needed**
  
```{r packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(RColorBrewer)
library(ggforce)

library(here)
sessionInfo()
```


*Directories*
  
```{r directories}
repoDir <- here::here()

# master data
dataDir <- '/Users/dein121/local/data_nonRepo/2025_irrigationAndYields/final_data_repo'
dataName <- 'pointSample_combined_fewerCDL_20240205.rds'

figDir <- paste0(repoDir,'/Figures_manuscript')
```


# Load and filter
Clean master data by:

* remove spurious classifications (eg, annual status differs from overarching irrigation type)
* remove the "existing irrigation" class
* focus on irrigation year of adoption -> condense each point into 1 observation
 
```{r loadMaster}
# load data, extract corn
master <- readRDS(paste0(dataDir,'/',dataName))


master1 <- master %>% 
  filter(irr_type != 'existing')
nrow(master1)

# check point type numbers
masterSlice <- master %>% 
  group_by(geom_id) %>%
  slice(1) %>%
  ungroup()

table(masterSlice$irr_type)
```

## new irrigation
reduce to 1 record each point
```{r process_new}
new <- master1 %>% 
  filter(irr_type == 'new')

# main record: adoption year
new_one<-new %>%
  filter(year == adoptionYear)

# find dominant previous crop
dominantCropBefore <- new %>%
  filter(year < adoptionYear) %>%
  group_by(geom_id, cdl_class) %>%
  summarize(frequency = n()) %>%
  ungroup() %>%
  group_by(geom_id) %>%
  filter(frequency == max(frequency)) %>%
  ungroup()
table(dominantCropBefore$cdl_class)

# find dominant subsequent crop
dominantCropAfter <- new %>%
  filter(year >= adoptionYear) %>%
  group_by(geom_id, cdl_class) %>%
  summarize(frequency = n()) %>%
  ungroup() %>%
  group_by(geom_id) %>%
  filter(frequency == max(frequency)) %>%
  ungroup()
table(dominantCropAfter$cdl_class)


```


## static variables only

```{r getStatic}
static0 <- master1 %>%
  group_by(geom_id) %>%
  slice(1) %>%
  ungroup()
table(static0$irr_type)

# drop annual vars
static <- static0[,1:44]

# recod irr type
static_all <- static %>%
  mutate(irr_status = case_when(irr_type == 'new' ~ 1,
                                irr_type == 'never - random' ~ 0,
                                irr_type == 'never - counter' ~ 0)) %>%
  mutate(pr_junJul_norm = pr_jul_norm_mm + pr_jun_norm_mm)

# only counterfactuals
static_counter <- static_all %>%
  filter(irr_type != 'never - random')
```




# logistic regression

## using all negatives
includes never - random and never-counterfactual

```{r logit}

# df with z scores
static_scaled <- static_all
cols <- c('dist_toIrr','dist_toStream','wtd_zell', 'ksat', 
          'vpd_jul_norm_hPa','pr_junJul_norm')

# change sign to go from "distance" to "proximity"
colsToChangeSign <- c('dist_toIrr','dist_toStream','wtd_zell')
static_scaled[cols] <- scale(static_scaled[cols])  
static_scaled[colsToChangeSign] <- static_scaled[colsToChangeSign]*-1


static_scaled <- static_scaled %>%
  rename(proxIrr = dist_toIrr,
         proxStream = dist_toStream,
         proxWT = wtd_zell)


model1 <- glm(irr_status ~ proxIrr + proxStream + proxWT +
                ksat + vpd_jul_norm_hPa + pr_junJul_norm  ,
              data = static_scaled, family = 'binomial')

summary(model1)
exp(coef(model1))

# get CI and translate coeffs + CI to odds ratios
conf95 <- confint(model1)
conf95_exp <- exp(conf95) %>%
  as.data.frame() %>%
  mutate(variable = row.names(.)) %>%
  rename(CI_lower = `2.5 %`,
         CI_higher = `97.5 %`)




modelCoeffs <- data.frame(variable = names(exp(coef(model1))),
                          coefficient = exp(coef(model1))) %>%
               as_tibble() %>%
               filter(variable != '(Intercept)')%>%
  left_join(conf95_exp)

modelCoeffs

1/.03 # field close to irrigaiton is 33 x more likely to be irrigated
      # than a field one standard deviation further away

# manipulate based on =/- effect
modelCoeffs2 <- modelCoeffs %>%
  mutate(coeff2 = coefficient - 1,
         CI_lower2 = CI_lower -1,
         CI_higher2 = CI_higher -1) 
```


# ms plot: Fig 4
plot it so they appear against the "1" odds ratio

```{r ms_plot}
varPalette = c(rep('#4393c3',3),
               rep('#d6604d',2),
                   '#fdae61')

modelCoeffs2$variable2 <- factor(c('Proximity to Irrigation',
                                   'Proximity to Stream',
                                   'Proximity to Water Table',
                                   'Soil Ksat',
                                   'July VPD',
                                   'June-July Ppt'),
                                levels = c("Proximity to Irrigation",
                                             'Proximity to Stream',
                                          'Proximity to Water Table',
                                                 'July VPD',
                                                  'June-July Ppt',
                                           'Soil Ksat'))

p_odds <- ggplot(modelCoeffs2,
       aes(x = variable2, y = coeff2, fill = variable2)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(ymin = CI_lower2, ymax = CI_higher2), width = .3) +
  coord_cartesian(ylim = c(-0.25, .75)) +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  scale_fill_manual(values = varPalette) +
  scale_y_continuous(name = "Odds Ratio", 
                   breaks = c(-.25, 0, .25,.5,.75),
                   labels = c("0.75",'1.00','1.25','1.50','1.75')) +
  xlab('') +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = 'none')
p_odds


ggsave(filename = paste0(figDir,'/Fig4_likelihood_oddsRatio.png'),
       plot =p_odds, dpi = 600,
       device = 'png' , width = 2.5, height = 3.5, units = 'in')

ggsave(filename = paste0(figDir,'/Fig4_likelihood_oddsRatio.pdf'),
       plot =p_odds, dpi = 600,
       device = 'pdf' , width = 2.5, height = 3.5, units = 'in')
```

