---
title: "Irrigated Area Summary Stats"
author: "jill deines/sophie baur"
date: '2024-02-05'
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
---
  
Goal: Assess historic changes in irrigated area based on NASS stats

Also includes:

* Generation of Figure 1c
* Generation of Figure 2a
* Calculating statistics for intro and study area background


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path='../figure/06.00_Fig_irrTrends/',
                      cache = FALSE)
```

**R Packages Needed**
  
```{r packages, warning=FALSE, message=FALSE}

library(dplyr)
library(ggplot2)
library(geofacet)
library(tidyr)
library(readr)
library(geofacet)
library(patchwork)
library(viridis)
library(RColorBrewer)

library(here)
sessionInfo()
```


*Directories*
  
```{r directories}
# nass data files
repoDir <- here::here()
nassDir <- paste0(repoDir,'/data/tabular/NASS')


nassFileName_raw_irrigated <- 'NASS_Census_IrrArea_CornBeltWide_1997-2022.csv'
nassFileName_raw_cropland <- 'NASS_Census_CroplandArea_CornBeltWide_1997-2022.csv'

# full 2017 census to extract irrigation by crop type
census2017raw_fn <- '/Users/dein121/local/data_nonRepo/2025_quickstats/qs.census2017.txt.gz'

# NASS Census irrigated area BY CROP (compiled in 00.30_checkNASS_irrigationByCrop.Rmd)
compiled_nassIrrDataName <- 'Compiled_CENSUS_2017_irrigated_byCropType_7states.csv'

# for production summary
nassFileName_raw_maize_NAME <- 'NASS_Survey_RAW_Maize_production_yields_harvestedArea_CONUS_1997-2023_20250323.csv'
nassFileName_raw_SOYBEAN_NAME <- 'NASS_Survey_RAW_Soybeans_production_yields_harvestedArea_CONUS_1997-2023_20250323.csv'


# output figures -------------
figDir <- paste0(repoDir,'/Figures_manuscript')
```


# Load 

## national irrigated area 
(from NASS Census via 00.20...Rmd)

```{r load1}
# load county data; also aggregate to state
irr_county <- read_csv(paste0(nassDir,'/',nassFileName_raw_irrigated ))
irr_state <- irr_county %>%
  group_by(state_alpha, state_name, year) %>%
  summarize(irrArea_ha = sum(irrArea_ha))

#define study area
study_area <- c("MN","WI","IA","IL","IN","MI","OH")

#subset for study area
irr_state_sa <- irr_state[irr_state$state_alpha %in% (study_area),]

#sum irrigated area for the study area states by year
irr_area_sa_year <- irr_state_sa %>%
  group_by(year) %>%
  summarize(irrArea_ha = sum(irrArea_ha))
```

## national cropland area

```{r load2}
#load in cropland dataset
crop_county <- read_csv(paste0(nassDir,'/',nassFileName_raw_cropland))

#merge dataframes
crop_irr_county <- merge(irr_county, crop_county, by = c("county_name","state_name","year"))

#group to state level
crop_irr_state <- crop_irr_county %>%
  group_by(state_alpha.x, state_name, year) %>%
  summarize(irrArea_ha = sum(irrArea_ha), 
            croplandArea_ha = sum(croplandArea_ha)) %>%
  ungroup()
crop_irr_state$irr_percent <- (crop_irr_state$irrArea_ha/crop_irr_state$croplandArea_ha)*100

```

## maize and soy production

```{r load3}
# load NASS
maize_prod00 <- read_csv(paste0(nassDir,'/',nassFileName_raw_maize_NAME))
soybeans_prod00 <- read_csv(paste0(nassDir,'/',nassFileName_raw_SOYBEAN_NAME))
```

## summarized irr/cropland by crop group
from 00.30...Rmd based on full 2017 census

```{r load4}
irr_byCrop <- read_csv(paste0(nassDir,'/',compiled_nassIrrDataName))
```


# MS fig: Fig 1c, trends by states

```{r}
palette_OkabeIto <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                      "#0072B2", "#D55E00", "#CC79A7", "#999999")
palette_OkabeIto_black_newYellow <- c("#E69F00", "#56B4E9", "#009E73", 
                                     # "#F5C710", 
                            "#0072B2", "#D55E00", "#CC79A7", "#000000")

pal_state <- c('#CC6677','#44AA99','#88CCEE','#888888','#AA4499','goldenrod2','#332288')

p_studyArea <- ggplot(irr_state_sa,
       aes(x = year, y = irrArea_ha/1000, group = state_alpha,
           color = state_alpha)) +
  geom_point() + geom_line() +
  #scale_color_viridis(discrete = T, option = 'turbo') +
  scale_color_manual(values = (palette_OkabeIto_black_newYellow)) +
 # facet_wrap(~state_name, nrow = 2) +
  ylab('') +
  theme_bw() + theme(legend.position = 'bottom',
                     legend.title = element_blank()) +
  ggtitle('Study Area by State')

# total study area
p_studyAreaTot <- ggplot(irr_area_sa_year,
       aes(x = year, y = irrArea_ha/1000)) +
  geom_line() + geom_point() +
  ylab('') +
  theme_bw() + ggtitle('Study Area Total')


# rest of conus
p_conus <- ggplot(irr_area_sa_year,
       aes(x = year, y = irrArea_ha/1000)) +
  geom_line() + geom_point() +
  theme_bw() +
   ylab('Irrigated Area (1000 ha)') +
  ggtitle('CONUS (Non - Study Area)')

#combine
p_all <- p_conus + p_studyAreaTot + p_studyArea &
  xlab('Year' )  &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"),
        text = element_text(size = 10))
p_all


# three wide
ggsave(filename = paste0(figDir,'/Fig1c_NASS_trends_studyAreaTotals_3panel.png'),
       plot =p_all, dpi = 600,
       device = 'png' , width = 6.35, height = 3.15, units = 'in')

ggsave(filename = paste0(figDir,'/Fig1c_NASS_trends_studyAreaTotals_3panel.pdf'),
       plot =p_all, dpi = 600,
       device = 'pdf' , width = 6.35, height = 3.15, units = 'in')
```

# MS fig: Fig 2a, irr by type


```{r fig2a}

p_areaByCropBoth <- ggplot(irr_byCrop,
      aes(x = reorder(COMMODITY_DESC, -percent), y = percent, fill = type2 )) +
  geom_bar(stat = 'identity', position = 'dodge') +
  xlab('Commodity') + ylab('Percent of Total') +
  theme_bw() +
  scale_fill_manual(values = c('#E69F00','#0072B2')) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.title = element_blank(),
        legend.position = c(.7,.7),
        panel.grid.minor.y = element_blank())  
p_areaByCropBoth

ggsave(filename = paste0(figDir,'/Fig2a_NASS_AreaByCrop_Both.png'),
       plot =p_areaByCropBoth, dpi = 600,
       device = 'png' , width = 3.15, height = 3.5, units = 'in')

ggsave(filename = paste0(figDir,'/Fig2a_NASS_AreaByCrop_Both.pdf'),
       plot =p_areaByCropBoth, dpi = 600,
       device = 'pdf' , width = 3.15, height = 3.5, units = 'in')
```

# MS statistics
## irrigation as a percent of cropped area

Summarize percent of relevant areas

```{r}
#define study area
study_area <- c("MN","WI","IA","IL","IN","MI","OH")

crop_irr_state %>%
  filter(state_alpha.x %in% study_area) %>%
  group_by(year) %>%
  summarize(irrArea_ha = sum(irrArea_ha),
            croplandArea_ha = sum(croplandArea_ha)) %>%
  mutate(irr_percent = irrArea_ha/croplandArea_ha*100)

```


quantify the increase in irrigated area for the full study region as a whole, as well as by state

compare with regions outside the study area

```{r}
#grab irrigated area for years 1997, 2002, 2007, and 2022
irr_area_1997_sa <- irr_area_sa_year$irrArea_ha[irr_area_sa_year$year == 1997]
irr_area_2022_sa <- irr_area_sa_year$irrArea_ha[irr_area_sa_year$year == 2022]

#calculate percent change from 1997-2022
study_area_change_1997 <- ((irr_area_2022_sa - irr_area_1997_sa)/(irr_area_1997_sa))*100 


paste0("From 1997 to 2022, the irrigated area within the study area increased by ", study_area_change_1997, "%.")

```


2) Quantify the change in irrigated area in other regions. (all states within CONUS except the study area)

```{r}
#subset contiguous US (without study area)
irr_state_nsa <- irr_state[!(irr_state$state_alpha %in% (study_area) | irr_state$state_alpha %in% c("AK","HI")),]

#sum the irrigated area for the contiguous US without study area, by year
irr_area_nsa_year <- irr_state_nsa %>%
  group_by(year) %>%
  summarize(irrArea_ha = sum(irrArea_ha))

#grab irrigated area for years 2022 and 1997
irr_area_2022_nsa <- irr_area_nsa_year$irrArea_ha[irr_area_nsa_year$year == 2022]
irr_area_1997_nsa <- irr_area_nsa_year$irrArea_ha[irr_area_nsa_year$year == 1997]

#calculate percent change from 1997-2002
non_study_area_change <- ((irr_area_2022_nsa - irr_area_1997_nsa)/(irr_area_1997_nsa))*100 
  
paste0("From 1997 to 2022, the irrigated area within the contiguous US, excluding the study area, increased by ", non_study_area_change, "%.")

```

## production stats

### maize
```{r maize_production}
#define study area
study_area <- c("MN","WI","IA","IL","IN","MI","OH")

# state production by year
maize_prod0 <- maize_prod00 %>% 
  filter(short_desc == 'CORN, GRAIN - PRODUCTION, MEASURED IN BU') %>%
  dplyr::select(c(state_alpha,year,Value, county_code)) %>%
  group_by(state_alpha, year) %>%
  summarize(production_bu = sum(Value)) %>%
  ungroup()

# maize study period mean
maize_prod <- maize_prod0 %>%
  filter(year >= 1997 & year <= 2017) %>%
  group_by(state_alpha) %>%
  summarize(production_bu = mean(production_bu))

# percent of maize production in  study area
maize_prod_sums <- maize_prod %>%
  mutate(isStudyArea = case_when(state_alpha %in% study_area ~ 1,
                            !(state_alpha %in% study_area) ~0)) %>%
  group_by(isStudyArea) %>%
  summarize(production_bu = sum(production_bu)) %>%
  tidyr::spread(., key = isStudyArea, value = production_bu) %>%
  mutate(pecentW =  `1` /( `0`+`1`) * 100)
maize_prod_sums
```


### soybeans
```{r soybeans_production}
#define study area
study_area <- c("MN","WI","IA","IL","IN","MI","OH")


# state production by year
soybeans_prod0 <- soybeans_prod00 %>% 
  filter(short_desc == 'SOYBEANS - PRODUCTION, MEASURED IN BU') %>%
  dplyr::select(c(state_alpha,year,Value, county_code)) %>%
  group_by(state_alpha, year) %>%
  summarize(production_bu = sum(Value)) %>%
  ungroup()

# soybeans study period mean
soybeans_prod <- soybeans_prod0 %>%
  filter(year >= 1997 & year <= 2017) %>%
  group_by(state_alpha) %>%
  summarize(production_bu = mean(production_bu))

# percent of soybeans production in  study area
soybeans_prod_sums <- soybeans_prod %>%
  mutate(isStudyArea = case_when(state_alpha %in% study_area ~ 1,
                            !(state_alpha %in% study_area) ~0)) %>%
  group_by(isStudyArea) %>%
  summarize(production_bu = sum(production_bu)) %>%
  tidyr::spread(., key = isStudyArea, value = production_bu) %>%
  mutate(pecentW =  `1` /( `0`+`1`) * 100)
soybeans_prod_sums
```









